### 2018-05-27 4_impl.tex
Cleanup finale up to unrolling.

The \ytree{} is compiled into the low-level event-based program representation called \textit{\xgraph{}}.
The mathematical structure of event-flow graph was discussed in Section~\ref{ch:wmm:model}.
The nodes of the graph are events, and the edges are basic relations: the control-flow relation \po{} and the data-flow relations \co{} and \rf{}.
Hereinafter, we denote the \xgraph{} with only control-flow edges as \xgraph[CF], the \xgraph{} with only data-flow edges as \xgraph[CF].
The complete \xgraph{} is $\xgraph[CF+DF] = \xgraph[CF] \; \cup \; \xgraph[DF]$.
The UML diagram on Figure~\ref{fig:class-diagrams:XEntity-interfaces} represents the hierarchy of main interfaces of the X-abstraction level.
Internally, the graph is represented by an adjacency matrix (to be exact, by multiple adjacency matrices that store edges of different kinds, see more details in Section~\ref{ch:impl:model:xgraph}).

@@

\paragraph{Memory units}
\label{ch:impl:model:xgraph:mem}

@@

\paragraph{Events}
\label{ch:impl:model:xgraph:evt}

@@

  \1 \texttt{XControlFlowEvent} \\
    The control-flow event indicates a non-linear jump in the code. We distinguish two kinds of control-flow events: 
    \2 \texttt{XJumpEvent} that performs no computation and no data operation, it can be safely removed from the model as an optimisation; and
    \2 \texttt{XMethodCallEvent} that models the function call with the \texttt{fastcall} calling convention (passing arguments in registers).
    The function call also implements the \texttt{XLocalMemoryUnit} since it represents the computed value returned by the function call.
    Currently, \porthos[2] only supports invocations of the method calls registered in its knowledge base (see invocation hooking mechanism described in Section~\ref{ch:impl:proc:x-compiler}).
    For interpreting known function invocations, \porthos[2] must perform three steps: bind actual arguments to the formal parameters (treated as temporary local memory units), push the called \texttt{XMethodCallEvent} onto the call stack, and the perform the control-flow jump to the function body.
    The call stack must be bounded by the user-defined parameter; once the stack is full, the interpretation should continue without jumping to the body of the invoked function (such cases must be properly logged).
    Each \texttt{return} statement must create the assignment of the function call event that is on the top of call stack.
    Note, this approach will work for recursion as well, and the recursive calls together with loops must be unrolled up the user-defined boundary (see Section~\ref{ch:impl:proc:x-unroll}).
    \\
    If the function has not been resolved, it can be safely assumed to be a no-operation function (with proper logging).
    In other words, we suppose that the knowledge base of \porthos[2] is complete and the tool can resolve all memory-operation functions and fence instructions.
    Although this can affect the completeness of the analysis, it is safe to make such an assumption as the user can check the log file and manually analyse the semantics of an unresolved call and add it to the knowledge base if necessary.

@@

\paragraph{Edges}
\label{ch:impl:model:xgraph:edges}
As the graph is represented by an adjacency matrix, its edges are stored in immutable hash-maps.
We distinguish the following kinds of edges:

@@

\paragraph{Graph invariants}
\label{ch:impl:model:xgraph:invariants}
Once being constructed, the graph must conform the following requirements (see also Section~\ref{ch:wmm:model:events}):

\begin{enumerate}[noitemsep]
\item the graph must have a single source with no ingoing edges, and two sinks of different kinds without outgoing edges,
\item the graph must be connected,
\item each node of the graph can have either one or two direct control-flow successors,
\item only nodes of type \texttt{XComputationEvent} can have two direct control-flow successors,
%\item each node except the source node must have at least one parent
\item a \co{}-edge connects two writes, an \rf{}-relation edge connects a write and a read,
\item all write-event nodes except initial write-nodes must have exactly one \co-predecessor, and
\item all write-event nodes except final write-nodes must have exactly one \co-successor.
%TODO: \item \textbf{TODO: more?}
\end{enumerate}

%\paragraph{Graph builder} % TODO: in code: remove graph hierarchy by combining unrolled and normal. Add a bool 'UnrolledStatus' -- thus we can keep genericity
%\label{ch:impl:model:xgraph:builder}
%TODO: code : NOTE: 'FlowGraph<N extends FlowGraphNode>' may be just 'FlowGraph<N>'

%\paragraph{Graph example} %(in 5_eval.tex)
%\label{ch:impl:model:xgraph:example}

%hashcode, uniqueness

%\lstinputlisting{inc/XEventVisitor-cleaned.java}

\subsubsection{W-model}
\label{ch:impl:model:wmodel}
%TODO: almost not implemented !

The \textit{\wmodel{}} represents a recursive AST of \textit{computations over relations} and assertion expressions defined by the memory model.
The atomic elements of \wmodel{} are the basic relations (\po{}, \rf{} and \co{}; see Section~\ref{ch:wmm:model:relations}) and sets of events ($\mathbb{R}$, $\mathbb{W}$, $\mathbb{IW}$, etc.; see Section~\ref{ch:wmm:model:events}).
The expressions of \wmodel{} are unary (such as complement, transitive closure, etc.) and binary operations (such as union, intersection, etc.) over relations or sets of events; see Section~\ref{ch:wmm:cat}.
For the sake of transparency, each element of a \wmodel{} contains the origin location as the coordinates of the string in the model file. %TODO: implement origin for w-model

%todo: enumerate elements -- after refactoring>

%todo: rename everywhere assertions <- axioms

@@

%TODO: not implemented at all !

The \textit{\zformula{}} representation is a wrapper for an SMT-formula used as an additional abstraction level used to increase the transparency of the architecture, simplify the debugging process and ease the support of different SMT-solvers.
As all other internal representations, the \zformula{} is implemented as an immutable data structure.

The Z-abstraction level models the the logical formulas (definitions and assertions) that are put on the assertion stack of the SMT-solver.
Generally, a \zformula{} represents the S-expression-based syntax of the SMT-LIB language~\cite{smt-lib}. % the semantics of SMT-LIB formulas defined in~\cite[Chapter 5]{smt-lib}.
Currently, the \zformula{} supports definitions and assertions over \textit{variables} and binary and unary \textit{expressions}.
However, in future it may be extended to support \textit{function symbols} and \textit{binders} (existential and universal quantification, pattern matching and functional type construction).

All expressions of a \zformula{} are \textit{typed} (or \textit{sorted}, in terms of SMT-LIB standard).
Basically, we distinguish \textit{boolean} and \textit{numeric} (integer, bitvector, real) expressions.
The typing of a \zformula{} is necessary for checking the basic validity of its expressions and converting it to a typed SMT-LIB formula.
For the sake of transparency, each element of a \zformula{} contains the origin location as the reference to the X-element that produced current Z-element. %TODO: implement origin for z-formula

%todo: enumerate elements -- after refactoring>

@@

%todo: after all, list HERE which modules (classes, managers) are being initialised during the pre-compilation. I'ts good to have the dependencies graph on moduls (operational semantics:) )

This section describes program units that construct, transform and analyse internal representations described before. % steps of construction of the program representations and transformations applied to them.
We assign unique numbers to the processing units, which can checked in Figure~\ref{fig:arch} (within gears depicted in the top-left corner of each processing unit).
%Further in current Section, the subsection titles have a number in parentheses, which correspond to a unique number of the  processing unit being discussed.
%These numbers are pictured within the gear in the top-left corner of each processing unit in Figure~\ref{fig:arch}.
%he numbers in parentheses correspond the numbers of processing units pictured within gears in Figure~\ref{fig:arch}.

The construction of the \xgraph{} is performed in three stages.
First, the \ytree{} is compiled to a \textit{cyclic} control-flow event-based graph \xgraph[CF].
Then, this graph is unrolled to an \textit{acyclic} control-flow event-based graph \xgraphU[CF].
After that, the compiler is able to perform the data-flow analysis and produce the full event-based graph \xgraphU[CF+DF], which remains to be \textit{CF-acyclic} (no cycles among control-flow edges).

Most data structures are processed by units that implement the \textit{visitor} pattern~\cite{palsberg1998essence}.
This is a behavioural pattern that separates the program logic from the object implementation by specifying handling methods for each element of the object.
The general structure of the visitor pattern is illustrated by the pseudo-java code in Figure~\ref{fig:visitor}.
The visitor pattern performs the double-dispatching, a mechanism for decreasing the cohesion between the DTO (the visitee) and the processor class (the visitor): the visitee implements the \textit{accepting} method that gets the visitor as an argument and invokes its \textit{visiting} method with itself as an argument.
Thus, the method call resolution is performed statically at compile-time without any overhead at run-time.
%To visit an instance \lstinline{x} by a visitor \lstinline{v}, the program should call the accept-method of \lstinline{x} and pass \lstinline{v} as an argument.
In our implementation, the visitor (not the visitee) cares about the continuation of traversing its children.

\begin{figure}[h]
\centering
%\begin{multicols*}{2}
\begin{minipage}[t]{.55\textwidth}
\begin{lstlisting}[language=java]
interface Element {
  <T> T accept(Visitor<T> visitor);
  ...
}

class AnElement implements Element {
  @Override
  <T> T accept(Visitor<T> visitor) {
    return visitor.visit(this);
  }
  ...
}
\end{lstlisting}
%[* \vfill\null \columnbreak *]
\end{minipage}% <---------------- Note the use of "%"
\begin{minipage}[t]{.45\textwidth}
\begin{lstlisting}[language=java]
class Visitor<T> {
  T visit(AnElement e) {
    // visiting logic
    ...
    // continue recursively
    e.getChild().accept(this);
  }
  T visit(AnOtherElement e) {
    ...
  }
  ...
}
\end{lstlisting}
%\end{multicols*}
\end{minipage}
\caption{Illustration of the visitor pattern}
\label{fig:visitor}
\end{figure}

@@

For traversing graph data structures, we mostly follow the \textit{iterator} pattern: the special object \texttt{Iterator}, that has access to elements of the data structure, iterates over them in some order (for example, in topological order as it is implemented for \xgraph{}).
%Since in \porthos{} we usually do not need to browse an object elements in different order, the iterator creation is performed by the data structure itself.
%However, in contrast to the visitor pattern, the visitor of a plain data structure should be used with an \textit{iterator} that has access to the elements of the object and passes them to the visitor.
The construction of immutable data structures is performed by units that follow the \textit{builder} pattern.
A builder is a mutable object that contains methods "filling" it by the elements that will constitute the result complex object.
The methods contain basic validity checks of the elements.
Once a builder has been built, it cannot be modified.

% todo: test pre-/post-fix operations //+think how current tool will behave if there's more than one post-operation //and pre-operation

@@

Both input-language and input-model parsers are implemented via \tool{ANTLR} parser generator.
%In Figure~\ref{fig:arch}, the input parsers are denoted as (1) and (2) processing units.
Details on the input program language grammar are discussed in Section~\ref{ch:impl:input}.
Currently, \porthos[2] does not consider preprocessor instructions (it ignores them).
For implementing the inter-procedural mode of \porthos{}, it is crucial to support inclusion of header files (the \texttt{\#include} preprocessor directive).
Next step would be implementing the support of macros and conditional compilation directives, that are used often in C code.
As preprocessor statements may appear in arbitrary place of a program, the preprocessor must be a stateful processing unit that reads the token stream and dynamically instrument the program by interpreting directives and expanding macros.
%The \tool{ANTLR} grammar for the C preprocessor may be found 
%more macros (stateful) : https://gcc.gnu.org/onlinedocs/cpp/Macros.html

The input memory model language \cat{} is discussed in Section~\ref{ch:wmm:cat}.
The \tool{ANTLR} grammar for \cat{} language was extracted from the parser used by \tool{herd} tool%
%
\footnote{The \tool{herd} project repository: \url{https://github.com/herd/herdtools7}} %
%
written in OCaml.
%TODO: upload the grammar to github
%The input memory model 
%TODO: add reference to the grammar file in github repository OR include a short version as an appendix}

%todo: maybe: move something from the input language chapter here

@@

%TODO: <not implemented yet>
The \wmodel{} representation is constructed by the stateless visitor \texttt{Cat2WmodelConverterVisitor} from the ANTLR syntax tree \texttt{CatParser.MainContext} of the memory-model defined in \cat{} language.
%In Figure~\ref{fig:arch}, the \wmodel{} constructor is denoted as the (3) processing unit.
Currently, the visitor supports the small subset of \cat{} language, which constitutes only non-functional declarative expressions of the language as the support for functional-style expressions requires implementing the full OCaml interpreter.
However, some most commonly used functional-style expressions may be supported by mapping them syntactically to corresponding W-elements by the \wmodel{} constructor.

Following the principle of transparency, the \wmodel{} constructor aborts its work with the \texttt{NotSupportedException} if met the unrecognised syntax construction (in contrast to the \porthos[1] approach in which the \texttt{null} value was produced in all exceptional states).

%todo: file tree of w-model to appendix -- after code fixing>

@@

The \ytree{} is constructed by the stateless visitor \texttt{C2YtreeConverterVisitor} from the ANTLR syntax tree \texttt{C11Parser.MainContext} of the C language.
As the \wmodel{} constructor, the \ytree{} constructor aborts its work with the \texttt{NotSupportedException} once it meets an unsupported syntax construction.
%TODO: Rename YConverterException <- YParserException
A syntax exception \texttt{YConverterException} is thrown if the converted syntax tree contains semantic errors that prevent it to be converted to the \ytree{}.

@@

%TODO: not supported yet

\paragraph{The label resolution}
\label{ch:impl:proc:x-pre-compiler:label}

@@

The label resolution is necessary for establishing links to labelled statements. %a process of linkage the text labels referenced in code to the places of their declaration.
In C, labelled statements are declared via the colon-syntax `\texttt{<label>\;:\;<statement>}', and the labels are referenced by the jump-statement `\texttt{goto\;<label>}'.
The label resolution algorithm traverses the \ytree{} and collects all declared labels into a map that points a label to the labeled statement. %TODO: code: remove temp label generation, it's stupid. put null.
This information is used during compilation to set up unconditional jumps.

\paragraph{Type analysis}
\label{ch:impl:proc:x-pre-compiler:type}

C language has a static (resolved at compile-time) manifest (all types are declared explicitly) type system.
Comparing to languages that use type inference, the type analysis of a C program constitutes a simple propagating the type information (obtained from variables declarations) to all expressions. % according the rules of propagation. For example, 
Being carried at Y- and X- representation levels, the type is converted to a~Z-type at the stage of the Z-formula encoding (see Section~\ref{ch:impl:proc:z-encoder}).

%TODO: YOU REALLY NEED TO THINK MORE ABOUT TYPES ON Y, X, Z LEVELS + reconsider this paragraph (and others about types)
Currently, \porthos[2] handles only the primitive C types (such as \texttt{int}, \texttt{char}, \texttt{float}, etc.), which is modelled as an enumeration \texttt{XType}. %TODO: model it.
The array dimension is stored at the Y-level as an integer number (currently this information is not used as \porthos[2] does not support arrays, although this information can be used at the stage of pre-compilation for converting array elements to separate variables). %the array operations are converted to ).
In addition, \porthos[2] has a built-in extensible database storing the semantics of non-primitive C types stored by the \texttt{X2ZTypeConverter}, %TODO: Or XTypeHookManager?
which can be requested for converting an X-type to a~Z-type. %TODO: implement this.
%TODO: describe the XType (before) + implement this clearly! + definitely get rid of bit32

%TODO: <not implemented>
The type analysis algorithm should consider the type aliases supported by C language (defined by the \texttt{typedef} instruction).
For resolving the type aliases, the precompiler should make an extra traverse of the \ytree{} before the pre-compilation stage and build up a symbol map.

%TODO: <not implemented>
The type analysis also includes the process of resolving semantics of function invocations.
The result is stored in a map of invocation expression to the resolved function signature, so that the compiler can decide either to jump to the function body and interpret it, or to hook the invocation if the semantics of the function is resolved.
However, due to lack of polymorphism and function overloading mechanism inherent to OOP languages, the function resolution algorithm may set up the mapping only for the function name and thus may neglect analysis of types of the arguments, we build the mapping for the full function signature so that the type analysis preprocessing unit may be used for analysing the C++ code if needed.

%TODOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOOO
%The X-type construction: \textbf{TODO} //code of class YType2TypeConverter (todo: rename: it's not conversion, it's a type construction, from the y-level name+qualifiers+
%- on Z level, type is one-pointer-level
%- The typification algorithm: propagation rules for each of Y
%- as vector array as needed.
%%
%Y: quantifiers+specifiers+typeName
%X: ?.. Z? TODO: rename+rework XType->XZType
%Z: Int, Real, Boolean, BitVector[N]

\paragraph{Variable kind analysis}
\label{ch:impl:proc:x-pre-compiler:var}

@@

%\vspace{1em}
The following types of variables are detected as \textit{global variables}:
%TODO: not implemented
\begin{itemize}
\item a variable was declared as a pointer;
\item a variable whose address was accessed by any process;
\item a variable that was declared as a parameter of the process function; and
%TODO: say sth about KERNEL ANALYSIS MODE BEFORE !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
\item a variable that was exported by the \texttt{extern} keyword (in the kernel-analysis mode, the functions \texttt{EXPORT\_SYMBOL} and \texttt{EXPORT\_SYMBOL\_GPL} also export symbols for dynamic linking). %https://lwn.net/Articles/154602/
\end{itemize}

@@

The X-compiler is the main component that transforms the recursive \ytree{} data structure to the plain \xgraph{} representation.
It is a complex processing unit; Figure~\ref{fig:compiler} illustrates the relationship between the principal components of the X-compiler in the UML language.

@@

The main class representing the X-compiler is \texttt{Y2XConverter}. %TODO: draw it on Figure.
It receives as input the \ytree{}, the memory model kind and user settings (for instance, the interpreter mode defining the set of invocation hooks enabled during the analysis run).
The \texttt{Y2XConverter} creates an instance of the stateless visitor \texttt{Y2XConverterVisitor} that traverses the \ytree{} while invoking the stateful interpreter \texttt{XInterpreter}.
The \texttt{XInterpreter} carries the \xgraph{}-builder and provides \textit{action} methods for changing its state and thus the filling in the builder.
The interpreter requests the managers such as \texttt{XMemoryManager}, \texttt{XHookManager}, \texttt{XTypeManager} for additional information.

We distinguish three types of processes (each implementing the \texttt{XInterpreter} interface):
\begin{itemize}[noitemsep,topsep=0pt]
\item the prelude process \texttt{XPreludeInterpreter} (to be executed before all other processes) that allows only declaration of local or shared variables, computations and memory operations;
\item the postlude process \texttt{XPostludeInterpreter} (to be executed after all other processes) that allows only declaration of local variables, memory operations, computations and assertions;
\item the process \texttt{XProcessInterpreter} (to be executed in parallel) that allows all X-compiler interface operations except declaration of shared variables %(that should be declared during pre-compilation stage or during compilation of the prelude process)
and program assertions (i.e., it allows declarations of local variables, computations, barriers, unconditional jumps, non-linear branching statements, and method calls).
\end{itemize}

As the prelude process can declare shared variables, its compilation must go first.
%TODO: think about sorting processes at the pre-compilation stage
The first event of each process must be the \texttt{XEntryEvent} and the last events must be events of type \texttt{XExitEvent}.

\paragraph{Memory manager}
\label{ch:impl:proc:x-compiler:mem}

@@

\paragraph{Invocation hook manager}
\label{ch:impl:proc:x-compiler:hooking}

%TODO: cite again here : alglave2018frightening + other Kernel references

%MethCallE : bind parameters with values (load them to temp registers) + remember the return-register
%reutrn event= assign the return-register + simple jump

>>>>

The \ytree{} is compiled into the low-level event-based program representation called \textit{\xgraph{}}.
The mathematical structure of event-flow graph was discussed in Section~\ref{ch:wmm:model}.
The nodes of the graph are events, and the edges are basic relations: the control-flow relation \po{} and the data-flow relations \co{} and \rf{}.
Hereinafter, we denote the \xgraph{} with only control-flow edges as \xgraph[CF], the \xgraph{} with only data-flow edges as \xgraph[CF].
The complete \xgraph{} is $\xgraph[CF+DF] = \xgraph[CF] \; \cup \; \xgraph[DF]$.
The UML diagram on Figure~\ref{fig:class-diagrams:XEntity-interfaces} represents the hierarchy of main interfaces of the X-abstraction level.
The \textit{full} list of classes that constitute the \xgraph{} hierarchy can be found in Appendix~\ref{apx:trees}.
Internally, the graph is represented by an adjacency matrix (to be exact, by multiple adjacency matrices that store edges of different kinds, see more details in Section~\ref{ch:impl:model:xgraph}).

@@

\paragraph{Memory units}
%\label{ch:impl:model:xgraph:mem}

@@

\paragraph{Events}
%\label{ch:impl:model:xgraph:evt}

@@

  \1 \texttt{XControlFlowEvent} \\
    The control-flow event indicates a non-linear jump in the code. We distinguish two kinds of control-flow events: 
    \2 \texttt{XJumpEvent} that performs no computation and no data operation, it can be safely removed from the model as an optimisation; and
    \2 \texttt{XFunctionCallEvent} that models function invocations.
    The function call also implements the \texttt{XLocalMemoryUnit} since it represents the computed value returned by the function call.
    Currently, \porthos[2] only supports invocations of the method calls registered in its knowledge base (see invocation hooking mechanism described in Section~\ref{ch:impl:proc:x-compiler}).
    For interpreting known function invocations, \porthos[2] must perform three steps: bind actual arguments to the formal parameters (treated as temporary registers as in the \texttt{fastcall} calling convention), push the called \texttt{XFunctionCallEvent} onto the call stack, and the perform the control-flow jump to the function body.
    The call stack must be bounded by the user-defined parameter; once the stack is full, the interpretation should continue without jumping to the body of the invoked function (such cases must be properly logged).
    Each \texttt{return} statement must create the assignment of the function call event that is on the top of call stack.
    Note, this approach will work for recursion as well, and the recursive calls together with loops must be unrolled up the user-defined boundary (see Section~\ref{ch:impl:proc:x-unroll}).
    \\
    If the function has not been resolved, it can be safely assumed to be a no-operation function (with proper logging).
    In other words, we suppose that the knowledge base of \porthos[2] is complete and the tool can resolve all memory-operation functions and fence instructions.
    Although this can affect the completeness of the analysis, it is safe to make such an assumption as the user can check the log file and manually analyse the semantics of an unresolved call and add it to the knowledge base if necessary.

@@

\paragraph{Edges}
%\label{ch:impl:model:xgraph:edges}
As the graph is represented by an adjacency matrix, its edges are stored in immutable hash-maps.
We distinguish the following kinds of edges:

@@

\paragraph{Graph invariants}
%\label{ch:impl:model:xgraph:invariants}
Once being constructed, the graph must conform the following requirements (see also Section~\ref{ch:wmm:model:events}):

\begin{enumerate}[noitemsep]
\item the graph must have a single source with no ingoing edges, and two sinks of different kinds without outgoing edges,
\item the graph must be connected,
\item each node of the graph can have either one or two direct control-flow successors,
\item only nodes of type \texttt{XComputationEvent} can have two direct control-flow successors,
%\item each node except the source node must have at least one parent
\item a \co{}-edge connects two writes, an \rf{}-relation edge connects a write and a read,
\item all write-event nodes except initial write-nodes must have exactly one \co-predecessor, and
\item all write-event nodes except final write-nodes must have exactly one \co-successor.
\end{enumerate}

%\paragraph{Graph builder} % TODO: in code: remove graph hierarchy by combining unrolled and normal. Add a bool 'UnrolledStatus' -- thus we can keep genericity
%TO-DO: code : NOTE: 'FlowGraph<N extends FlowGraphNode>' may be just 'FlowGraph<N>'

\subsubsection{W-model}
\label{ch:impl:model:wmodel}

The \textit{\wmodel{}} represents a recursive AST of \textit{computation over relations} and \textit{assertion expressions} defined by the memory model.
The atomic elements of \wmodel{} are the basic relations (\po{}, \rf{} and \co{}; see Section~\ref{ch:wmm:model:relations}) and sets of events ($\mathbb{R}$, $\mathbb{W}$, $\mathbb{IW}$, etc.; see Section~\ref{ch:wmm:model:events}).
The expressions of \wmodel{} are unary (such as complement, transitive closure, etc.) and binary operations (such as union, intersection, etc.) over relations or sets of events; see Section~\ref{ch:wmm:cat}.
Each element of \wmodel{} implements the interface \texttt{WModel}, and for the sake of transparency contains the origin location in the model file. %TODO: implement origin for w-model
As the \wmodel{} hierarchy almost completely follows the mathematical structure discussed in Sections~\ref{ch:wmm:cat} and~\ref{ch:enc:bmc:wmm}, we skip the details of its implementation.

@@

The \textit{\zformula{}} representation is supposed to be a wrapper over an SMT-formula used as an additional abstraction level to increase the transparency of the architecture, simplify the debugging process and ease the support of different SMT-solvers.
However, currently \porthos[2] skips this internal representation and both the program and memory models directly into the SMT-formula via Java API offered by the Z3 solver.
%As all other internal representations, the \zformula{} should be implemented as an immutable data structure.

The Z-abstraction level should model the logical formulas (definitions and assertions) that are put on the assertion stack of the SMT-solver.
Generally, a \zformula{} should represent the S-expression-based syntax of the SMT-LIB language~\cite{smt-lib}.
Currently, the \zformula{} is supposed to represent definitions and assertions over variables and binary and unary expressions.
However, for the purpose of improving the encoding scheme, it may be extended to support function symbols and binders (existential and universal quantification, pattern matching and functional type construction).

All expressions of a \zformula{} must be \textit{typed} (or \textit{sorted}, in terms of SMT-LIB standard).
Basically, the Z-type must support boolean and numeric (integer, bitvector, real) expressions.
The typing of a \zformula{} is necessary for checking the basic validity of its expressions and converting it to a typed SMT-formula.
For the sake of transparency, each element of a \zformula{} should contains the origin location as the reference to the X-element that produced current Z-element.

@@

This section describes program units that construct, transform and analyse internal representations described above.
%not anymore %Each processing unit is assigned with a unique number, which can checked in Figure~\ref{fig:arch} (within gears depicted in the top-left corner of each processing unit).

The construction of the \xgraph{} is performed in three stages.
Firstly, the \ytree{} is compiled to a \textit{cyclic} control-flow event-based graph \xgraph[CF].
Then, this graph is unrolled to an \textit{acyclic} control-flow event-based graph \xgraphU[CF].
After that, the compiler is able to perform the data-flow analysis over the \xgraphU[CF] and produce the full event-based graph \xgraphU[CF+DF], which remains to be \textit{CF-acyclic} (no cycles among control-flow edges).

Most data structures are processed by units that implement the \textit{visitor} pattern~\cite{palsberg1998essence}.
This is a behavioural pattern that separates the program logic from the object implementation by specifying handling methods for each element of the object.
The general structure of the visitor pattern is illustrated by the pseudo-java code in Figure~\ref{fig:visitor}.
The visitor pattern performs the double-dispatching, a mechanism for decreasing the cohesion between the DTO (the visitee) and the processor class (the visitor): the visitee implements the \textit{accepting} method that gets the visitor as an argument and invokes its \textit{visiting} method with itself as an argument.
Thus, the method call resolution is performed statically at compile-time without any overhead at run-time.
In our implementation, the visitor (not the visitee) cares about the continuation of traversing its children.

\begin{figure}[h]
\centering
\begin{minipage}[t]{.55\textwidth}
\begin{lstlisting}[language=java]
interface Element {
  <T> T accept(Visitor<T> visitor);
  ...
}

class AnElement implements Element {
  @Override
  <T> T accept(Visitor<T> visitor) {
    return visitor.visit(this);
  }
  ...
}
\end{lstlisting}
\end{minipage}
\begin{minipage}[t]{.45\textwidth}
\begin{lstlisting}[language=java]
class Visitor<T> {
  T visit(AnElement e) {
    // visiting logic
    ...
    // continue recursively
    e.getChild().accept(this);
  }
  T visit(AnOtherElement e) {
    ...
  }
  ...
}
\end{lstlisting}
\end{minipage}
\caption{Illustration of the visitor pattern}
\label{fig:visitor}
\end{figure}

@@

For traversing plain data structures (such as graphs), we mostly follow the \textit{iterator} pattern: the special object \texttt{Iterator}, that has access to elements of the data structure, iterates over them in some order (for example, in topological order as it is implemented for \xgraph{}).
The construction of immutable data structures is performed by units that follow the \textit{builder} pattern.
A builder is a mutable object that contains methods for "filling" it with the elements that will constitute the result complex object.
The methods contain basic validity checks of the elements.
Once a builder has been built, it cannot be modified.

@@

As it was discussed in Section~\ref{ch:impl:input}, both input-language and input-model parsers are implemented via \tool{ANTLR} parser generator.
Currently, \porthos[2] does not consider C preprocessor instructions (it ignores them).
For implementing the inter-procedural mode of \porthos{}, it is crucial to support inclusion of header files (the \texttt{\#include} preprocessor directive).
Next step would be implementing the support of macros and conditional compilation directives, that are used often in C code.
As preprocessor statements may appear in arbitrary place of a program, the preprocessor must be a stateful processing unit that reads the token stream and dynamically instrument the program by interpreting directives and expanding macros.

The input memory model language \cat{} is discussed in Section~\ref{ch:wmm:cat}.
The \tool{ANTLR} grammar for \cat{} language was extracted from the parser used by \tool{herd} tool%
%
\footnote{The \tool{herd} project repository: \url{https://github.com/herd/herdtools7}} %
%
written in OCaml.
%TODO: upload the grammar to github

@@

%TODO: <not implemented yet>
The \wmodel{} representation is constructed by the stateless visitor \texttt{Cat2WmodelConverterVisitor} from the ANTLR syntax tree \texttt{CatParser.MainContext} of the memory-model defined in \cat{} language.
Currently, the visitor supports the small subset of \cat{} language, which constitutes only non-functional declarative expressions of the language as the support for functional-style expressions requires implementing the full interpreter for OCaml-like language.
However, some most commonly used functional-style expressions may be supported by mapping them syntactically to corresponding W-elements directly in the \wmodel{} constructor.

Following the principle of transparency, the \wmodel{} constructor aborts its work with the \texttt{NotSupportedException} if met the unsupported syntax construction (in contrast to the \porthos[1] approach in which the \texttt{null} value was produced in all exceptional states).

%todo: file tree of w-model to appendix -- after implementing + more discussion here

@@

The \ytree{} is constructed by the stateless visitor \texttt{C2YtreeConverterVisitor} from the ANTLR syntax tree \texttt{C11Parser.MainContext} of the C language.
The \ytree{} constructor aborts its work with the \texttt{YParserNotImplementedException} once it meets an unsupported syntax construction.
A syntax exception \texttt{YParserException} is thrown if the converted syntax tree contains semantic errors that prevent it to be converted to the \ytree{}.

@@

@@

\paragraph{The label resolution}
%\label{ch:impl:proc:x-pre-compiler:label}

The label resolution is necessary for establishing links to labelled statements.
In C, labelled statements are declared via the colon-syntax `\texttt{<label>\;:\;<statement>}', and the labels are referenced by the jump-statement `\texttt{goto\;<label>}'.
The label resolution algorithm traverses the \ytree{} and collects all declared labels into a map that points a label to the labelled statement.
This information is used during compilation to set up unconditional jumps.

\paragraph{Type analysis}
%\label{ch:impl:proc:x-pre-compiler:type}

C language has a static (resolved at compile-time) manifest (all types are declared explicitly) type system.
Comparing to languages that use type inference, the type analysis of a C program constitutes a simple propagating the type information (obtained from variables declarations) to all expressions.
Being carried at Y- and X- representation levels, the type is converted to a~Z-type at the stage of the Z-formula encoding (see Section~\ref{ch:impl:proc:z-encoder}).

Currently, \porthos[2] handles only the primitive C types (such as \texttt{int}, \texttt{char}, \texttt{float}, etc.), which are modelled on the X-level by the enumeration \texttt{XType}.
%TODO: describe the XType (before) + implement this clearly! + definitely get rid of bit32
%The array dimension is stored at the Y-level as an integer number (currently this information is not used as \porthos[2] does not support arrays, although this information can be used at the stage of pre-compilation for converting array elements to separate variables). %the array operations are converted to ).
%In addition, \porthos[2] has a built-in extensible database storing the semantics of non-primitive C types stored by the \texttt{X2ZTypeConverter},
%which can be requested for converting an X-type to a~Z-type. %TODO: implement this.
The type analysis algorithm should consider the type aliases supported by C language (defined by the \texttt{typedef} instruction), which is not implemented yet.
For resolving the type aliases, the precompiler should make an extra traverse of the \ytree{} before the pre-compilation stage and build up a symbol map.

%<not needed below>
%The type analysis also includes the process of resolving semantics of function invocations.
%The result is stored in a map of invocation expression to the resolved function signature, so that the compiler can decide either to jump to the function body and interpret it, or to hook the invocation if the semantics of the function is resolved.
%However, due to lack of polymorphism and function overloading mechanism inherent to OOP languages, the function resolution algorithm may set up the mapping only for the function name and thus may neglect analysis of types of the arguments, we build the mapping for the full function signature so that the type analysis processing unit may be used for analysing the C++ code if needed.

\paragraph{Variable kind analysis}
%\label{ch:impl:proc:x-pre-compiler:var}

@@

The following types of variables are detected as \textit{global variables}:
%TODO: <not implemented yet>
\begin{itemize}[noitemsep,topsep=0pt]
\item a variable was declared as a pointer;
\item a variable whose address was accessed by any process;
\item a variable declared as a parameter of the process function; and
\item a variable exported by the \texttt{extern} keyword (in the kernel-analysis mode, the functions \texttt{EXPORT\_SYMBOL} and \texttt{EXPORT\_SYMBOL\_GPL} also export symbols for dynamic linking~\cite{argawal16kernel}).
\end{itemize}

@@

The \textit{X-compiler} is the main component that transforms the recursive \ytree{} data structure to the plain \xgraph{} representation.
It is a complex processing unit; Figure~\ref{fig:compiler} illustrates the relationship between main components of the X-compiler in the UML language.

@@

The main class representing the X-compiler is \texttt{Y2XConverter}. %TODO: draw it on Figure.
It receives as input the \ytree{}, the memory model kind and user settings (for instance, the interpreter mode defining the set of invocation hooks enabled during the analysis run).
The \texttt{Y2XConverter} creates an instance of the stateless visitor \texttt{Y2XConverterVisitor} that traverses the \ytree{} while invoking the stateful interpreter \texttt{XInterpreter}.
The \texttt{XInterpreter} carries the \xgraph{}-builder and provides \textit{action} methods for changing its state and thus the filling in the builder.
The interpreter has additional modules \texttt{XMemoryManager}, \texttt{XHookManager} and \texttt{XTypeManager} that provide specialised functionality.

We distinguish three types of interpreters (each implementing the \texttt{XInterpreter} interface):
\begin{itemize}[noitemsep,topsep=0pt]
\item the \textit{program} interpreter \texttt{XProgramInterpreter} that is the compositional element that dispatches calls to currently maintained process interpreter;
\item the \textit{prelude} process \texttt{XPreludeInterpreter} (to be executed before all other processes) that allows only declaration of local or shared variables, computations and memory operations;
\item the \textit{postlude} process \texttt{XPostludeInterpreter} (to be executed after all other processes) that allows only declaration of local variables, memory operations, computations and assertions;
\item the \textit{process} \texttt{XProcessInterpreter} (to be executed in parallel) that allows all X-compiler interface operations except declaration of shared variables
and program assertions (i.e., it allows declarations of local variables, computations, barriers, unconditional jumps, non-linear branching statements, and method calls).
\end{itemize}

As the prelude process can declare shared variables, its compilation must go first.
%TODO: sort processes at the pre-compilation stage !
The first event of each process must be the \texttt{XEntryEvent} and the last events must be events of type \texttt{XExitEvent}.

\paragraph{Memory manager}
%\label{ch:impl:proc:x-compiler:mem}

@@

\paragraph{Invocation hooking manager}
%\label{ch:impl:proc:x-compiler:hooking}
